---
title: "Exploratory Data Analysis"
format: 
  html:
    embed-resources: true
html-math-method: 
  method: katex
editor: source
execute:
  warning: false
---

This notebook documents the exploratory data analysis phase for this project. The results of this have informed our data-cleaning procedures.

## Explore the data

### gamma-scale

### all taxa, gamma-scale

Load the data for all the taxa at the gamma-scale:

```{r}
# load the data
gam_all <- readr::read_delim(here::here("Data-raw", "gamma-all-raw.csv"), delim = ";")
head(gam_all)

```

Explore the structure of the data

```{r}
# check the names
names(gam_all)
```
```{r}
# check the dimensions
dim(gam_all)
```

```{r}
# check the basic variable types
dplyr::glimpse(gam_all)
```

#### Summary statistics

```{r}
# calculate basic summary statistics for each variable using summary()
summary(gam_all)
```

#### Data visualisation

Check the distribution of each numeric variable:

```{r}
# plot histograms of each variable

```


### active dispersers, gamma-scale

Load the data for all the active dispersers at the gamma-scale:

```{r}
# load the data
gam_active <- readr::read_delim(here::here("Data-raw/gamma-active-raw.csv"), ";")
head(gam_active)
```

Explore the structure of the data

```{r}
# check the names
names(gam_active )
```

### passive dispersers, gamma-scale

Load the data for all the passive dispersers at the gamma-scale:

```{r}
# load the data
gam_passive <- readr::read_delim(here::here("Data-raw/gamma-passive-raw.csv"), ";")
head(gam_passive)
```

Explore the structure of the data

```{r}
# check the names
names(gam_active )
```

### alpha-scale

### all taxa, alpha-scale

Load the data for all taxa at the alpha-scale:

```{r}
# load the data
alpha_all <- readr::read_delim(here::here("Data-raw/alpha-all-raw.csv"))
head(alpha_all)
```

Explore the structure of the data

```{r}
# check the names
names(alpha_all)
```

Check the data for potential outliers:

```{r}
# check depth of the pools
summary(alpha_all)
```

### active dispersers, alpha-scale

Load the data for all the active dispersers at the alpha-scale:

```{r}
# load the data
alpha_active <- readr::read_delim(here::here("Data-raw/alpha-active-raw.csv"), ";")
head(alpha_active)
```

Explore the structure of the data

```{r}
# check the names
names(alpha_active)
```

```{r}
# how many rows?
nrow(alpha_active)
```

### passive dispersers, alpha-scale

Load the data for all the passive dispersers at the alpha-scale:

```{r}
# load the data
alpha_passive <- readr::read_delim(here::here("Data-raw/alpha-passive-raw.csv"), ";")
head(alpha_passive)
```

Explore the structure of the data

```{r}
# check the names
names(alpha_passive)
```






