---
title: "Species accumulation curves"
format: 
  html:
    embed-resources: true
html-math-method: 
  method: katex
editor: source
execute:
  warning: false
---

Here we use the raw Use the community abundance data to examine the species accumulation curves for each inselberg for the supplementary material.

We start by loading the relevant libraries:

```{r}
# load relevant libraries
library(dplyr)
library(readr)
library(vegan)
library(ggplot2)
library(ggpubr)
```

Next, we load the raw community abundance data and remove the single inselberg from Columbia as the size fraction that was used in the plankton nets was too different from the other inselbergs.

```{r}
# load the relevant data
sac_dat <- readr::read_csv(here::here("Data-raw/pool-community-abundance-data.csv"))

```

Next, we split the community data by inselberg so that the data becomes a list where each list element represents the data from a different inselberg. We then write a function (SACplot) that implements the species accumulation curve via rarefaction and plots this curve.

```{r}
# split the data by location
sac_list <- split(sac_dat, sac_dat$Inselberg)

# function to plot the species accumulation curve
SACplot <- function(data, location) {
  
  # get the community without the identifiers
  sp <- data.frame(sapply(data[,-c(1:2)], as.numeric))
  
  # round abundances
  sp <- round(sp, digits = 0)
  
  # retain columns with at least one taxa
  sp_pres <- sp[, colSums(sp) > 0]
  
  # perform the rarefaction with 10000 permutations
  sac_curve <- specaccum(sp_pres, method = "rarefaction", permutations = 10000)
  
  # get data.frame for the accumulation curve
  sac_curve_df <- data.frame(cbind(sac_curve[["sites"]],sac_curve[["richness"]],sac_curve[["sd"]])) 
  
  # rename the columns
  colnames(sac_curve_df) <- c("sites","richness","sd")
  
  # make sd ribbon intervals
  sac_curve_df[["low_sd"]] <- sac_curve_df[["richness"]]-sac_curve_df[["sd"]]
  sac_curve_df[["high_sd"]] <- sac_curve_df[["richness"]]+sac_curve_df[["sd"]]
  
  p1 <- 
    ggplot(data = sac_curve_df)+
    geom_line(mapping = aes(x = sites, y = richness), size = 0.65) +
    geom_ribbon(mapping = aes(x = sites, ymin = low_sd, ymax = high_sd), alpha = 0.1) +
    labs(x = "", y = "") +
    annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -1,
             label = location, parse = TRUE) +
    theme_classic()
  
  p1
  
}
```

Using this function, we create a species accumulation curve for each inselberg:

```{r}
# for each inselberg, make an accumlation curve
figs2.1 <- vector("list", length = length(sac_list))
for(i in 1:length(sac_list)) {
  
  # get an SAC plot for each inselberg
  figs2.1[[i]] <- SACplot(data = sac_list[[i]], location = sac_list[[i]]$Inselberg[1])
  
}

# check the figures
plot(figs2.1[[1]])
```

We then combine this list of plots into a single plot for the supplementary material:

```{r}
# appendix: fig. s2.1

# plot the data
figs2.1 <- 
  ggarrange(plotlist = figs2.1, ncol = 4, nrow = 3,
            font.label = list(size = 11, face = "plain"),
            common.legend = TRUE, legend = "bottom")

# annotate the data
figs2.1 <- annotate_figure(figs2.1,
                           left = text_grob("Rarefied richness", rot = 90, vjust = 1, size = 14),
                           bottom = text_grob("Number of Rock pools", vjust = -1, size = 14))

figs2.1
```

Export the combined figure:

```{r}
# export the figure
ggsave(here::here("Figures/fig_s2.1.pdf"), figs2.1, units = "cm", width = 20, height = 13)
```


